\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-15 paper.qmd
more changes and suggestions on methods section and application

@@

The \gD{} hexagon grid is defined by the number of bins in each direction $(b_1, b_2)$, giving total number of bins as $b = b_1 \times b_2$, and a unique hexagon id, $h = 1, \dots, b$. Each hexagon, $H_h$ is uniquely described by centroid, $C_{h}^{(2)} = (c_{h1}, c_{h2})$.  (Note that, we explain the algorithm for constructing the grid and binning the data here because, surprisingly, there is no readily available implementation that could produce everything needed for this project.)

To make the grid, a buffer parameter ($q$) is set as a proportion of the plot space. By default this is 0.1. A value should be chosen to provide at buffer of a full hexagon width ($a_1$) and height ($a_2$) for the grid to extend beyond the data, in all directions. The lower left position where the grid starts is defined as $(s_1, s_2)$, and corresponds to the centroid of the lowest left hexagon. This must be smaller than the minimum data value. This is computed as one buffer unit, $q$ below the minimum data value, by default. The user provides the preferred number of bins in the horizontal direction, $b_1$, and the algorithm computes the appropriate number of bins in the vertical direction, $b_2$.

@@

Delaunay triangulation [@lee1980] is used to connect neighboring centroids, which is needed to preserve neighborhood information when the model is lifted into \pD{}.

@@

@@

@@

@@

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "Reproduction of plot published in Chen et al. (2023) showing a $2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. The parameter settings, beyond the defaults, are n_neighbors = 30 and min_dist = 0.3. The question is whether this is an accurate representation of structure in the high-dimensional data."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H

>>>>

XXX The points on the "theoretical model" should be equidistant along the S. They are still to bunched in some spots and too far apart in others.

@@

Although there are several implementations of hexagon binning (REF), and a published paper (REF), surprisingly, none has sufficient detail or components that produce everything needed for this project. So we described the process used here. @fig-hex-param illustrates the notation used.

The \gD{} hexagon grid is defined by its bin centroids. Each hexagon, $H_h$ ($h = 1, \dots, b$) is uniquely described by centroid, $C_{h}^{(2)} = (c_{h1}, c_{h2})$. The number of bins in each direction is denoted as $(b_1, b_2)$, with  $b = b_1 \times b_2$ being the total number of bins. We expect the user to provide just $b_1$ and we calculate $b_2$ using the NLDR ratio, to compute the grid.

To ensure that the grid covers the range of data values a buffer parameter ($q$) is set as a proportion of the range. By default,  $q=0.1$. The buffer should be extending a full hexagon width ($a_1$) and height ($a_2$) beyond the data, in all directions. The lower left position where the grid starts is defined as $(s_1, s_2)$, and corresponds to the centroid of the lowest left hexagon, $C_{1}^{(2)} = (c_{11}, c_{12})$. This must be smaller than the minimum data value. Because it is one buffer unit, $q$ below the minimum data values, $s_1 = -q$ and $s_2 = -r_2*q$.

@@

Delaunay triangulation [@lee1980] (XXXREF what R package are you using?) is used to connect neighboring centroids. The edges are needed to preserve neighborhood information when the model is lifted into \pD{}.

@@

When shapes are non-linear in the NLDR layout, some edges could be long. It is unlikely

@@

XXX You need to have consistent colouring. In langevitour the points corresponding to observations are purple. They should be purple in the \gD{} view too.

XXX Should there be a second long edge removal? You can compute the edges in \pD{} now, and if there are long edges maybe these should be pruned, and we re-plot the \gD{} view.

@@

XXX Prediction should have a jitter option, to spread the predicted points out fully in the hexagons.

@@

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "Reproduction of plot published in Chen et al. (2023) showing a $2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. The parameter settings, beyond the defaults, are 30 nearest neighbors and minimum distance equal to 0.3. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H

### 2024-07-15 paper.tex
more changes and suggestions on methods section and application

@@

The \gD{} hexagon grid is defined by the number of bins in each
direction \((b_1, b_2)\), giving total number of bins as
\(b = b_1 \times b_2\), and a unique hexagon id, \(h = 1, \dots, b\).
Each hexagon, \(H_h\) is uniquely described by centroid,
\(C_{h}^{(2)} = (c_{h1}, c_{h2})\). (Note that, we explain the algorithm
for constructing the grid and binning the data here because,
surprisingly, there is no readily available implementation that could
produce everything needed for this project.)

To make the grid, a buffer parameter (\(q\)) is set as a proportion of
the plot space. By default this is 0.1. A value should be chosen to
provide at buffer of a full hexagon width (\(a_1\)) and height (\(a_2\))
for the grid to extend beyond the data, in all directions. The lower
left position where the grid starts is defined as \((s_1, s_2)\), and
corresponds to the centroid of the lowest left hexagon. This must be
smaller than the minimum data value. This is computed as one buffer
unit, \(q\) below the minimum data value, by default. The user provides
the preferred number of bins in the horizontal direction, \(b_1\), and
the algorithm computes the appropriate number of bins in the vertical
direction, \(b_2\).

@@

Delaunay triangulation \citep{lee1980} is used to connect neighboring
centroids, which is needed to preserve neighborhood information when the
model is lifted into \pD{}.

@@

\caption{\label{fig-umap-author}Reproduction of plot published in Chen
et al.~(2023) showing a \(2\text{-}D\) layout from UMAP applied for the
PBMC3k dataset. The parameter settings, beyond the defaults, are
n\_neighbors = 30 and min\_dist = 0.3. The question is whether this is
an accurate representation of structure in the high-dimensional data.}

>>>>

XXX The points on the ``theoretical model'' should be equidistant along
the S. They are still to bunched in some spots and too far apart in
others.

@@

Although there are several implementations of hexagon binning (REF), and
a published paper (REF), surprisingly, none has sufficient detail or
components that produce everything needed for this project. So we
described the process used here. Figure~\ref{fig-hex-param} illustrates
the notation used.

The \gD{} hexagon grid is defined by its bin centroids. Each hexagon,
\(H_h\) (\(h = 1, \dots, b\)) is uniquely described by centroid,
\(C_{h}^{(2)} = (c_{h1}, c_{h2})\). The number of bins in each direction
is denoted as \((b_1, b_2)\), with \(b = b_1 \times b_2\) being the
total number of bins. We expect the user to provide just \(b_1\) and we
calculate \(b_2\) using the NLDR ratio, to compute the grid.

To ensure that the grid covers the range of data values a buffer
parameter (\(q\)) is set as a proportion of the range. By default,
\(q=0.1\). The buffer should be extending a full hexagon width (\(a_1\))
and height (\(a_2\)) beyond the data, in all directions. The lower left
position where the grid starts is defined as \((s_1, s_2)\), and
corresponds to the centroid of the lowest left hexagon,
\(C_{1}^{(2)} = (c_{11}, c_{12})\). This must be smaller than the
minimum data value. Because it is one buffer unit, \(q\) below the
minimum data values, \(s_1 = -q\) and \(s_2 = -r_2*q\).

@@

Delaunay triangulation \citep{lee1980} (XXXREF what R package are you
using?) is used to connect neighboring centroids. The edges are needed
to preserve neighborhood information when the model is lifted into
\pD{}.

When shapes are non-linear in the NLDR layout, some edges could be long.
It is unlikely

@@

\caption{\label{fig-umap-author}Reproduction of plot published in Chen
et al.~(2023) showing a \(2\text{-}D\) layout from UMAP applied for the
PBMC3k dataset. The parameter settings, beyond the defaults, are 30
nearest neighbors and minimum distance equal to 0.3. We use our
model-in-the-data-space to assess whether this is an accurate
representation of structure present in the high-dimensional data, or if
it is misleading.}
